---
title: 'Assignment 2 Task 2: Parameter Estimation with Purrr'
author: "Matthieu Huy"
date: "2023-02-11"
output: html_document
---

```{r setup, echo = TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(tidymodels)
library(kableExtra)
library(here)
library(cowplot)
library(GGally)
library(ggtext)
library(AICcmodavg)
library(jtools)
library(Metrics)
```

This analysis uses data containing observations of lizard species, sex, length, and weight in the Jornada Basin, located in the Chihuahaun desert of New Mexico.

Citation:\
- Lightfoot, D. and W.G. Whitford. 2020. Lizard pitfall trap data from 11 NPP study locations at the Jornada Basin LTER site, 1989-2006 ver 37. Environmental Data Initiative. https://doi.org/10.6073/pasta/4a6e258fb49c31e222ecbbcfd128967f (Accessed 2023-02-11).


Metadata:\
- https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-jrn.210007001.37


For this analysis, I will use non linear least squares to estimate parameters of a length to weight model for lizard populations in New Mexico. 

### Read in and clean data 

```{r}
lizards_data <- read_csv(here("data/lizards.csv")) |>
  janitor::clean_names() |>
  rename(length = sv_length, species = spp)

lizard_species_code <- read_table(here("data/Lizardcodelist.txt"), col_names = FALSE) |>
  unite(scientific_name, X2, X3) |>
  unite(common_name, X4, X5, X6) |>
  select(code = X1, scientific_name, common_name) |>
  tail(19)
```

### 1) Fit a length to weight model of the following form to all lizards in your data frame\

$$
{W}=a*(SVL)^b
$$
### log transform the equattion to get:

$$
log(W) = log(a) + b*log(SVL)
$$

- W: weight
- SVL: snout-to-vent length
- a and b are parameters that need to be fitted

```{r}
#log transform the data for initial guess model
lizards_data <- lizards_data |>
  mutate(log_length = log(length), 
         log_weight = log(weight))

my_guess_model <- lm(log_weight ~ log_length, data = lizards_data)
coefs_tr <- exp(coefficients(my_guess_model)) #transformed coefficients


```

### NLS Model
```{r}
nls_model <- function(a, length, b){
  out=a*(length)^b
  }
lizards_nls = nls(weight ~ nls_model(a, length, b),
                 data = lizards_data,
                 start = list(a = exp(-8.25), b = 2.48),
                 trace = TRUE)

broom::tidy(lizards_nls) %>% 
  kable(caption = "Lizards NLS") %>% 
  kable_classic()
```

```{r}
# Make a prediction based on the model

lizards_predict<- lizards_data %>% 
  mutate(predict = predict(lizards_nls,newdata=.))

labels <- as_labeller(c("M" = "Male", "F" = "Female"))

ggplot(data = lizards_predict) +
  geom_point(aes(x = length, 
                 y = weight,
                 color = sex)) +
  geom_smooth(aes(x = length, y = predict),
              color='red') +
  facet_wrap(~sex,
             labeller = labels) +
  theme_light() +
  theme(legend.position = "none") +
  labs(x = "Length",
       y = "Weight")
#color code and theme and stuff

```
\
```{r}
#create subset of male Western Whiptail lizards
lizards_subset <- lizards_data |>
  filter(species %in% c("CNTI"),
         sex %in% c("M"))

subset_nls = nls(weight ~ nls_model(a, length, b),
                 data = lizards_subset,
                 start = list(a = exp(-8.25), b = 2.48),
                 trace = TRUE)

broom::tidy(subset_nls) %>% 
  kable(caption = "Western Whiptail Species Specific NLS") %>% 
  kable_classic()
```
\
```{r}

whiptail_predict<- lizards_subset %>% 
  mutate(predict = predict(subset_nls, newdata=.))

rmse_full_nls <- rmse(lizards_predict$weight, lizards_predict$predict)
rmse_whip_nls <- rmse(whiptail_predict$weight, whiptail_predict$predict)

ggplot() +
  geom_point(data = whiptail_predict,
             aes(x = length, 
                 y = weight,
                 color = sex),
             color = "darkcyan",
             alpha = 0.6) +
  geom_smooth(data = whiptail_predict,
              aes(x = length, y = predict),
              color='red') +
  geom_smooth(data = lizards_predict,
              aes(x = length, y = predict),
              color='darkcyan') +
  labs(x = "Length",
       y = "Weight",
       caption = "**RMSE is
       <span style ='color:red;'>1.563 using the full model</span> and
       <span style ='color:darkcyan;'>3.349 using the whiptail subset model.</span>**<br>
       Given that the model parameterized on the full dataset has a lower RMSE, it would be best to use this model to<br> predict weight given the snout-to-vent length of a whiptail lizard.") +
  theme_light() +
  theme(
        plot.caption = element_markdown(size = 10, hjust = 0, 
                                        lineheight = 1.2, margin = margin(10, 0, 0, -25)),
        axis.text = element_text(color = "black", size = 10),                 
        axis.title.x = element_text(color = "black", size = 12, face = "bold"),  
        axis.title.y = element_text(color = "black", size = 12, face = "bold",
                                    vjust = 2), 
        legend.position = "none",
        legend.text = element_text(size = 8))
```

